Multiresolution Neural Field Estimation

### 2011-03-02 Multiresolution_framework/ltx/manuscript.tex
edit to figure and intro

\section{Introduction}
\IEEEPARstart{T}{he} human brain has a multiresolution architecture, where spatial scales for information processing and transfer range from individual proteins in synapses, to neurons, to networks of neurons. The multiresolution cortical architecture poses major modeling challenges to efficiently describe the brain's dynamics. This paper introduces a multiresolution data-driven framework for neural field modeling, the multiresolution approximate integrodifference equation (MRAIDE), to address this challenge.

Neural field models typically describe the meso and macroscopic dynamics of the human brain, but are also descriptive of finer-scale neurodynamics. The ability to create patient-specific neural models will contribute to our knowledge base of poorly understood diseases such as epilepsy, and furthermore, may also enable the development of new treatment strategies. This is particularly poignant with the advent of new devices that use therapeutic electrical stimulation. Presently, stimulation strategies for devices operate in an open-loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists an enormous potential to improve the performance of these devices using systems theory. The design of a stimulation strategy using systems theory requires a model. Neural mass and neural field models are the ideal candidates for estimation and control algorithms. It is expected that parameters of the neural fields models will be patient specific and will therefore need to be inferred from data.

The first work describing data driven mesoscopic neural modeling~\cite{Valdes1999} used a neural mass model~\cite{LopesDaSilva1976,Zetterberg1978} to fit EEG data. This approach was extended to coupled neural masses (using another model~\cite{Jansen1995}) through a Bayesian estimation scheme dubbed Dynamic Causal Modeling (DCM)~\cite{David2003}. Following this work, data-driven modeling was extended to continuum field equations that explain the richer dynamics of spatiotemporal neural fields \cite{Galka2008,schiff2008kalman,Daunizeau2009}. Most recently a framework was developed where a finite element model of the neural field (via a global Galerkin projection) was formed, using a basis function decomposition, to transform the PDE neural field equations into a finite dimension system to facilitate efficient state and parameter estimation~\cite{Freestone2011}.

@@

The spatial cutoff frequency of the observed field was used to specify the level of decomposition required to represent the field using the wavelets. Fig.~\ref{fig:KernelAndFreqResponse} (b), shows the field spatial frequency and cuttoff of \dean{??}. Therefore, wavelets up to level $j=2$ (with the bandwidth $\approx[2.5,4]$ cycles/mm) can represent the significant spatial characteristics of the field. \cut{Note that wavelets due to their bandpass filter characteristics extract successively higher and higher frequency components by increasing $j$.}

@@

The histogram in \figurename{\ref{fig:ParametersHistogram}} demonstrates the performance of the parameter estimation. \dean{The results are good because the RMSE between the true and estimated kernels is $1.83$ (units?)... There is a bias of 7.4 \% in $\theta_1$ because..., which is not big deal.}
\begin{figure}[!h]
 \centering
 \includegraphics[scale=.9]{./Graph/ObservationFreqResponse.pdf}
 \caption{(a) Mexican-hat connectivity kernel. The kernel is a composite of two B-spline
functions (dashed lines), representing local excitation and lateral inhibition. (b) The average (over time) power in dB of the spatial frequency of the observations. The dashed line shows the cutoff frequency (-3 dB point).}
\label{fig:KernelAndFreqResponse} 
  \end{figure} 
\begin{figure}[!h] 
 \centering
 \includegraphics[scale=0.9]{./Graph/Field.pdf}
 \caption{Actual (solid line) and estimated (dashed line) neural fields at a time instant for different spatial resolutions. (a) $j=0$, $n_x=17$. (b) $j=1$, $n_x=33$. (c) $j=2$, $n_x=65$. (d) $j=3$, $n_x=131$. (e) $j=4$, $n_x=263$.}
 \label{fig:FieldEstimates}
 \end{figure} 
\begin{figure}[!h] 
 \centering
 \includegraphics[scale=1]{./Graph/Hist.pdf}
 \caption{(a) RMSE of the estimated field at different spatial scales. (b-c) Histograms of the excitatory and inhibitory connectivity basis function parameter estimates, $\theta_0$ and $\theta_1$, respectively. The solid lines show the actual parameter values and the dotted lines show the estimated parameters means.}
 \label{fig:ParametersHistogram}
 \end{figure}

>>>>

\section{Introduction}
\IEEEPARstart{T}{he} human brain has a multiresolution architecture, where spatial scales for information processing and transfer range from individual proteins in synapses, to neurons, to networks of neurons. The multiresolution cortical architecture poses major modeling challenges to efficiently describe the brain's dynamics. This paper introduces a multiresolution data-driven framework for neural field modeling, the multiresolution approximate integro-difference equation (MRAIDE), to address this challenge.

Neural field models describe the mass action of the central nervous system and are a critical link in our understanding of the biophysics of the EEG. They typically describe the macroscopic dynamics of the human brain, but are also descriptive of finer-scale neurodynamics. The ability to create patient-specific neural models will contribute to our knowledge base of poorly understood diseases such as epilepsy, and furthermore, may also enable the development of new treatment strategies. This is particularly poignant with the advent of new devices that use therapeutic electrical stimulation. Presently, stimulation strategies for devices operate in an open-loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists an enormous potential to improve the performance of these devices using systems theory, but first a suitable model is required. Neural mass and neural field models are the ideal candidates for estimation and control algorithms, due to their strong links with the underlying physiology and their parsimony. It is expected that parameters of the neural fields models will be patient specific and will therefore need to be inferred from data.

The first work describing data-driven mesoscopic neural modeling~\cite{Valdes1999} used a neural mass model~\cite{LopesDaSilva1976,Zetterberg1978} to fit EEG data. This approach was extended to coupled neural masses (using another model~\cite{Jansen1995}) through a Bayesian estimation scheme dubbed Dynamic Causal Modeling (DCM)~\cite{David2003}. Following this work, data-driven modeling was extended to continuum field equations that explain the richer dynamics of spatiotemporal neural fields \cite{Galka2008,schiff2008kalman,Daunizeau2009}. Most recently a framework was developed where a finite element model of the neural field (via a global Galerkin projection) was formed, using a basis function decomposition, to transform the PDE neural field equations into a finite dimension system to facilitate efficient state and parameter estimation~\cite{Freestone2011}.

@@

The spatial cutoff frequency of the observed field was used to specify the level of decomposition required to represent the field using the wavelets. Fig.~\ref{fig:KernelAndFreqResponse} (b), shows the field spatial frequency and cutoff of \dean{??}. Therefore, wavelets up to level $j=2$ (with the bandwidth $\approx[2.5,4]$ cycles/mm) can represent the significant spatial characteristics of the field. \cut{Note that wavelets due to their bandpass filter characteristics extract successively higher and higher frequency components by increasing $j$.}

@@

The histogram in \figurename{\ref{fig:ParametersHistogram}} demonstrates the performance of the parameter estimation. \dean{The results are good because the RMSE between the true and estimated kernels is $1.83$ (units?)... There is a bias of 7.4 \% in $\theta_1$ because..., which is not big deal.}
% \begin{figure}[!h]
%  \centering
%  \includegraphics[scale=.9]{./Graph/ObservationFreqResponse.pdf}
%  \caption{(a) Mexican-hat connectivity kernel. The kernel is a composite of two B-spline
% functions (dashed lines), representing local excitation and lateral inhibition. (b) The average (over time) power in dB of the spatial frequency of the observations. The dashed line shows the cutoff frequency (-3 dB point).}
% \label{fig:KernelAndFreqResponse} 
%   \end{figure} 
\begin{figure}[!h] 
 \centering
 \includegraphics[scale=0.9]{./Graph/Field.pdf}
 \caption{Actual (solid line) and estimated (dashed line) neural fields at a time instant for different spatial resolutions. (a) $j=0$, $n_x=17$, RMSE = . (b) $j=1$, $n_x=33$, RMSE = . (c) $j=2$, $n_x=65$, RMSE = . (d) $j=3$, $n_x=131$, RMSE = . (e) $j=4$, $n_x=263$, RMSE = .}
 \label{fig:FieldEstimates}
 \end{figure} 
% \begin{figure}[!h] 
%  \centering
%  \includegraphics[scale=1]{./Graph/Hist.pdf}
%  \caption{(a) RMSE of the estimated field at different spatial scales. (b-c) Histograms of the excitatory and inhibitory connectivity basis function parameter estimates, $\theta_0$ and $\theta_1$, respectively. The solid lines show the actual parameter values and the dotted lines show the estimated parameters means.}
%  \label{fig:ParametersHistogram}
%  \end{figure} 
\begin{figure}[!h] 
 \centering
 \includegraphics[scale=1]{./Graph/KernelEstimate.pdf}
 \caption{The actual connectivity kernel is shown the the black solid line. The estimated kernel and confidence interval is shown by the red line and red shaded region ($\pm2$ std). The 7 weighted kernel basis functions are shown by the black dashed lines.}
 \label{fig:KernelEstimate}
 \end{figure}
